
First, I would like to give my most sincere thanks to my supervisor Prof. Bruno C. d. S. Oliveira.
Six years ago, I was lucky to get a chance to exchange to HKU and
took two courses taught by Bruno,
and both of them are quite challenging and interesting at the same time.
Since then, I decided to explore more on programming languages and
therefore applied for Bruno's Ph.D. student.
It turns out to be a wise choice!
Regular meetings with him are usually quite informative, encouraging and fun.
Bruno is a very patient person;
he always provides professional suggestions on how to proceed
with the research when I get stuck or move in the wrong direction.
In short, I would never expect a better supervisor.

Collaborations in research have been unforgettable experiences.
Prof. Tom Schrijvers offered help on my type inference project,
giving me invaluable ideas and suggestions on both research, writing, and presentation.
This piece of work contributes to the most recognized publications during my Ph.D.,
which won the distinguished paper award in ICFP.
I also participated in projects of my colleagues Xuejing Huang and Yaoda Zhu,
which are both challenging and interesting. I learned a lot from you as well.
I would like to thank everyone in the group, for your enlightening seminars and discussions:
%
Tomas Tauber, Huang Li, Xuan Bi, Haoyuan Zhang, Yanlin Wang, Yanpeng Yang,
Weixin Zhang, Ningning Xie, 
%
Xuejing Huang, Yaoda Zhou, Baber Rehman, Mingqi Xue, Yaozhu Sun, Wenjia Ye, Xu Xue,
Chen Cui, Jinhao Tan.


Study and life in HKU were wonderful. Staff from various sections of the university
are kind and willing to help.
I would like to also thank Dr. Francois Pottier, Prof. Chuan Wu, Prof. Ravi Ramanathan
for your detailed review and precious feedbacks on my thesis.


I would also like to thank my friends who study or work in Hong Kong.
Discussions with Ge Bai were always insightful and fun.
Gatherings with high school classmates, including those who visit Hong Kong for a short time,
were memorable:
Yunze Deng, Bintao Sun, Lu Lu, Tong Yu, Junfeng Chen, Zhiyu Wan, Ran You,
Yiyao Xu, Yiming Hu, Hengyun Zhou, Jinglong Zhao.


Last but not least, my family is always supporting and encouraging.
My parents, grandparents, uncles and aunts give me warm advice through
my difficult times and on important decisions.
Above all, I would like to offer my grateful thanks to my wife Jingyu Zhao
for your love, tolerance and constant support.
You are the one who understands me the best, for being classmates in
middle school and high school, and we are both Ph.D. candidates in HKU.
Even when I feel desperate, the time spent with you is always delightful and encouraging.
Life would be much harder without you.

